---
title: "Getting Started with tidymatrix"
author: "tidymatrix package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tidymatrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates basic tidymatrix usage through practical examples including questionnaire data and gene expression analysis.

## Introduction

`tidymatrix` provides a unified interface for working with matrices and their associated metadata. The key idea is to keep your matrix data together with row and column annotations, making data manipulation more intuitive and less error-prone.

```{r load-packages}
library(tidymatrix)
library(dplyr)
```

## Example 1: Questionnaire Data

Let's start with survey data where we have:
- **Matrix**: survey responses (people × questions)
- **Row metadata**: demographics about respondents
- **Column metadata**: information about questions

### Creating the Data

```{r questionnaire-data}
set.seed(123)
n_people <- 100
n_questions <- 10

# Survey response matrix (Likert scale 1-5)
responses <- matrix(
  sample(1:5, n_people * n_questions, replace = TRUE),
  nrow = n_people,
  ncol = n_questions
)

# Row metadata: people demographics
people_data <- data.frame(
  person_id = 1:n_people,
  age = sample(18:80, n_people, replace = TRUE),
  gender = sample(c("M", "F", "Other"), n_people, replace = TRUE),
  education = sample(c("High School", "Bachelor", "Master", "PhD"),
                     n_people, replace = TRUE)
)

# Column metadata: question information
question_data <- data.frame(
  question_id = 1:n_questions,
  category = rep(c("Demographics", "Satisfaction", "Behavior"),
                 length.out = n_questions),
  question_text = paste("Question", 1:n_questions),
  required = sample(c(TRUE, FALSE), n_questions, replace = TRUE)
)

# Create tidymatrix object
tm <- tidymatrix(responses, people_data, question_data)
print(tm)
```

## Example 2: Working with Rows (People)

Use `activate(rows)` to work with row metadata (respondents):

### Filtering Respondents

```{r filter-rows}
# Filter to only people over 30
tm_filtered <- tm |>
  activate(rows) |>
  filter(age > 30)

print(tm_filtered)
```

### Adding Derived Columns

```{r mutate-rows}
# Add a new column: age group
tm_with_groups <- tm |>
  activate(rows) |>
  mutate(age_group = case_when(
    age < 30 ~ "Young",
    age < 50 ~ "Middle",
    TRUE ~ "Senior"
  ))

print(tm_with_groups)
```

### Selecting Metadata Columns

```{r select-rows}
# Select specific metadata columns
tm_selected <- tm |>
  activate(rows) |>
  select(person_id, age, gender)

print(tm_selected)
```

## Example 3: Working with Columns (Questions)

Use `activate(columns)` to work with column metadata (questions):

### Filtering Questions

```{r filter-columns}
# Filter to only required questions
tm_required <- tm |>
  activate(columns) |>
  filter(required == TRUE)

print(tm_required)
```

### Adding Question Labels

```{r mutate-columns}
# Add question numbers
tm_numbered <- tm |>
  activate(columns) |>
  mutate(q_number = paste0("Q", question_id))

print(tm_numbered)
```

## Example 4: Chaining Operations

You can chain operations across different components:

```{r chaining}
# Get satisfaction questions for people over 40
tm_complex <- tm |>
  activate(rows) |>
  filter(age > 40, gender == "F") |>
  activate(columns) |>
  filter(category == "Satisfaction") |>
  activate(rows) |>
  arrange(age)

print(tm_complex)
```

## Example 5: Gene Expression Data

A common use case is gene expression analysis:

```{r gene-expression}
n_genes <- 50
n_samples <- 20

# Expression matrix
expression <- matrix(
  rnorm(n_genes * n_samples, mean = 5, sd = 2),
  nrow = n_genes,
  ncol = n_samples
)

# Gene annotations
gene_data <- data.frame(
  gene_id = paste0("GENE", 1:n_genes),
  gene_symbol = paste0("SYM", 1:n_genes),
  chromosome = sample(1:22, n_genes, replace = TRUE),
  pathway = sample(c("Metabolism", "Signaling", "Transport"),
                   n_genes, replace = TRUE)
)

# Sample information
sample_data <- data.frame(
  sample_id = paste0("S", 1:n_samples),
  condition = rep(c("Control", "Treatment"), each = n_samples/2),
  batch = rep(1:4, each = n_samples/4),
  patient_id = rep(1:10, each = 2)
)

# Create gene expression tidymatrix
expr_tm <- tidymatrix(expression, gene_data, sample_data)

# Filter to metabolism genes and treatment samples
expr_filtered <- expr_tm |>
  activate(rows) |>
  filter(pathway == "Metabolism") |>
  activate(columns) |>
  filter(condition == "Treatment")

print(expr_filtered)
```

## Summary

The `tidymatrix` package provides:

- **Unified data structure**: Keep matrices and metadata together
- **dplyr verbs**: Use familiar `filter()`, `mutate()`, `select()`, etc.
- **Component activation**: Work with rows, columns, or the matrix itself
- **Pipe-friendly**: Chain operations naturally with `|>`

This makes matrix manipulation more intuitive and reduces the risk of misalignment between your data and metadata.
